---
title: "R Project to Help Visual Long-term Trends in Analytes"
subtitle: "Snake River at Flagg and Moose"
author: "Ben LaFrance, benjamin_lafrance@nps.gov"
date: "2025-01-30"
output: html_document
---

The following code blocks enable the generation of figures and key data/numbers that
GRYN uses for their annual water quality reports.

```{r LoadLibraries, message=FALSE, warning=FALSE, include=FALSE, results=FALSE}

# List of packages: checks whether they are installed
packages <- c('dataRetrieval', 'dplyr', 'ggplot2')
not_installed <- packages[!(packages %in% installed.packages()[ , "Package"])]

# Prompts installation if not installed
if(length(not_installed) > 0) {
  if (menu(c("Yes", "No"),
         title = paste("Are you sure you want to install package", 
                       not_installed)) == "1") {
    install.packages(not_installed)
  } else {
    print("Cancelling installation")
  }
}

# Load packages
invisible(lapply(packages, library, character.only = TRUE))
```

```{r DefineSites, include=FALSE}

# Store all relevant stream gage data
streamgages <- tribble(
  ~Site,  ~GageID,    ~StartDate,  ~EndDate, ~SiteName,                                  ~WQX_ID,
  "SNR1", "13010065", "1984-10-01", "",      "Snake River at Flagg, WY",                 "11NPSWRD_WQX-GRTE_SNR01",
  "SNR2", "13013650", "1995-10-01", "",      "Snake River at Moose, WY",                 "11NPSWRD_WQX-GRTE_SNR02",
  "YRCS", "06191500", "1911-01-01", "",      "Yellowstone River at Corwin Springs",      "11NPSWRD_WQX-YELL_YS549.7M",
  "LMR" , "06188000", "1923-10-01", "",      "Lamar River near Tower Junction, WY",      "11NPSWRD_WQX-YELL_LM000.5M",   
  "MDR" , "06037500", "1914-10-01", "",      "Madison River near West Yellowstone, MT",  "11NPSWRD_WQX-YELL_MDR",
  "SBC" , "06187915", "1999-10-01", "",      "Soda Butte Creek near Silver Gate, MT",    "11NPSWRD_WQX-YELL_SB015.7A",
  "SHR1", "06285100", "1966-10-01", "",      "Shoshone River near Lovell, WY",           "11NPSWRD_WQX-BICA_SHR1",
  "BHR1", "06287000", "1968-10-01", "",      "Bighorn River near St. Xavier, MT",        "11NPSWRD_WQX-BICA_BHR1",
  "BHR2", "06279500", "1929-04-01", "",      "Bighorn River at Kane, WY",                "11NPSWRD_WQX-BICA_BHR2"
  )


```

```{r DownloadAllDischargeDataFromUSGS}
# Function to get data for all GRYN associated Stream Gage Sites
# This will take 1-2 minutes, and output a few lines about how many days worth of data there are

# Update idea - do we need to do this for all sites, or just the site of interest?

get_gage_data <- function(site) {
  print(site$Site)
  Daily <- readNWISDaily(site$GageID, "00060", site$StartDate, site$EndDate) |>
    filter(grepl('A', Qualifier)) |>
    mutate(cfs = Q * 35.314666212661, Year = trunc(DecYear))
  
  # Return all of the discharge data for all of the sites
  return(tibble(Daily = list(Daily), 
                Site = site$Site, 
                SiteName = site$SiteName, 
                StartYear = year(ymd(site$StartDate)), 
                EndYear = if_else(site$EndDate == "", year(Sys.Date()), year(ymd(site$EndDate)))))
}
# Get data for all sites
gage_data <- bind_rows(lapply(split(streamgages, seq(nrow(streamgages))), get_gage_data))

```


```{r DefineAnalytes}

# Update idea: add the Field Parameters to this tribble? Wont work as mg/L anymore though so could be confusing
# Update idea: add the BICA and YELL parameters that don't overlap with GRTE

# Tribble of all the GRTE Analytes you can look up 
all_analytes <- tribble(
  ~Analyte,                         ~CharacteristicName,   ~ResultSampleFractionText,    ~WYChronicAquaticLifeCriterion,  ~WYHumanHeathCriterion,
  "Total Arsenic",                  "Arsenic",             "Total",                      0.150,                            0.010,
  "Dissolved Arsenic",              "Arsenic",             "Dissolved",                  0.150,                            0.010,
  "Total Calcium",                  "Calcium",             "Total",                      -9999,                            -9999,              
  "Dissolved Calcium",              "Calcium",             "Dissolved",                  -9999,                            -9999,
  "Total Copper",                   "Copper",              "Total",                      0.090,                            1.000,
  "Dissolved Copper",               "Copper",              "Dissolved",                  -9999,                            -9999,
  "Total Iron" ,                    "Iron",                "Total",                      1.000,                            0.300,
  "Dissolved Iron",                 "Iron",                "Dissolved",                  1.000,                            1.000,         
  "Total Magnesium",                "Magnesium",           "Total",                      -9999,                            -9999,
  "Dissolved Magnesium",            "Magnesium",           "Dissolved",                  -9999,                            -9999,      
  "Total Potassium",                "Potassium",           "Total",                      -9999,                            -9999,
  "Dissolved Potassium",            "Potassium",           "Dissolved",                  -9999,                            -9999,
  "Total Selenium",                 "Selenium",            "Total",                      0.005,                            0.050,
  "Dissolved Selenium",             "Selenium",            "Dissolved",                  0.005,                            0.050,
  "Total Sodium",                   "Sodium",              "Total",                      -9999,                            -9999,
  "Dissolved Sodium",               "Sodium",              "Dissolved",                  -9999,                            -9999,                  
  
  "Dissolved Chloride (Cl)",        "Chloride",            "Dissolved",                  230.0,                            -9999,
  "Sulfate (S04)",                  "Sulfate",             "Dissolved",                  -9999,                            -9999,
  "Total Phosphorus",               "Phosphorus",          "Total",                      -9999,                            -9999,                                        
  "Orthophosphate (Ortho-P)",       "Orthophosphate",      "Dissolved",                  -9999,                            -9999,                                         
  "Hardness/Alkalinity as CaCO3",   "Hardness, carbonate", "Total",                      -9999,                            -9999,                                         
 
  "Ammonia as Nitrogen (NH3 as N)", "Ammonia",             "Total",                      -9999,                            -9999,
  "Nitrate + Nitrite as Nitrogen", "Nitrate + Nitrite",    "Total",                      1.000,                            -9999, 
  "Total Suspended Solids (TSS)", "Total suspended solids", "Non-Filterable (Particle)", -9999,                            -9999       
)

# Ammonia criteria is pH and temperature dependent, see: https://www.epa.gov/sites/default/files/2014-12/documents/wy-chapter1.pdf
# In tables, we've been recording the column name as "Nitrate + Nitrite as Nitrogen (NO3 + NO2 as N)"
```


```{r FunctionGetAnalyteDataAndVisualizePerYear}

# Define a function to pull and process water quality data
visualize_analyte_data_per_year <- function(site_of_interest, analyte_name, result_fraction) {
  
  # Retrieve the WQX ID and Site Name from the streamgages tribble
  site_info <- streamgages %>% filter(Site == site_of_interest)
  
  if (nrow(site_info) == 0) {
    stop("Site not found")
  }
  
  site_id <- site_info$WQX_ID
  site_name <- site_info$SiteName

  # Read the data from the Water Quality Portal
  data <- dataRetrieval::readWQPdata(siteid = site_id, dataProfile = "resultPhysChem")

  # Filter the data using the specified analyte and result fraction
  filtered_data <- data |>
    filter(ActivityStartDate >= "2006-01-01" & ActivityStartDate <= "2023-12-31") |>
    filter(CharacteristicName == analyte_name & ResultSampleFractionText == result_fraction) |>
    select(ActivityStartDate, CharacteristicName, ResultSampleFractionText, 
           ResultMeasureValue, ResultMeasure.MeasureUnitCode, 
           DetectionQuantitationLimitMeasure.MeasureValue)

  # Replace NA values and prepare the data for plotting
  processed_data <- filtered_data |>
    mutate(ResultMeasureValue = as.numeric(coalesce(ResultMeasureValue, 
        as.character(DetectionQuantitationLimitMeasure.MeasureValue)))) |>
    group_by(ActivityStartDate) |>
    slice(1) |> 
    ungroup() |>
    mutate(Year = as.numeric(format(ActivityStartDate, "%Y")),
           SiteName = site_name)

  # Create the plot
  plot <- ggplot(processed_data, aes(x = Year, y = ResultMeasureValue, color = as.factor(Year))) +
    geom_line() +
    geom_point() +
    labs(title = paste("Annual", result_fraction, analyte_name, "Levels for", site_name),
         x = "Year",
         y = paste(analyte_name, "(", result_fraction, "mg/L)", sep = ""),
         color = "Year") +
    theme_minimal() +
    scale_color_discrete(type = rainbow(n = length(unique(processed_data$Year))))

  # Return both processed data and plot
  return(list(data = processed_data, plot = plot))
}

# Example usage for Total Arsenic
# SNR1_total_Arsenic <- visualize_analyte_data_per_year("SNR1", "Arsenic", "Total")

```


```{r PerYearVariationPlot}
# With the function provided above, you should be able to pick any site, see the examples below
# Allowed sites are provided in the "DefineSites" code block
# You should also be able to pick any analyte in the "DefineAnalytes" code block

SNR1_total_Arsenic <- visualize_analyte_data_per_year("SNR1", "Arsenic", "Total")
SNR2_dissolved_Arsenic <- visualize_analyte_data_per_year("SNR2", "Arsenic", "Dissolved")
MDR_total_phos <- visualize_analyte_data_per_year("MDR", "Phosphorus", "Total")

# Generate the graph
SNR1_total_Arsenic
SNR2_dissolved_Arsenic
MDR_total_phos
```
Next steps using the above code:
1. Visualize all analyte and location combinations available
2. Comb for outliers and either flag or correct in NPStoret before publishing 2024 data. 
3. Determine a priority list of analytes and sites for future trend analysis. Some analytes will be <RL too often to have a meaningful sample size.

The above code provides a way to look at the annual variation. Another way to visualize the data is to look at how the analytes vary with discharge throughout the year (and then stack each year on the same graph). That code is privded below:

```{r FunctionGetAnalyteDataAndVisualizeAnnualTrend}

# Define a function to pull and process water quality data
visualize_analyte_data_annual_trend <- function(site_of_interest, analyte_name, result_fraction) {
  
  # Retrieve the WQX ID and Site Name from the streamgages tribble
  site_info <- streamgages %>% filter(Site == site_of_interest)
  
  if (nrow(site_info) == 0) {
    stop("Site not found")
  }
  
  site_id <- site_info$WQX_ID
  site_name <- site_info$SiteName

  # Read the data from the Water Quality Portal
  data <- dataRetrieval::readWQPdata(siteid = site_id, dataProfile = "resultPhysChem")

  # Filter the data using the specified analyte and result fraction
  filtered_data <- data |>
    filter(ActivityStartDate >= "2006-01-01" & ActivityStartDate <= "2023-12-31") |>
    filter(CharacteristicName == analyte_name & ResultSampleFractionText == result_fraction) |>
    select(ActivityStartDate, CharacteristicName, ResultSampleFractionText, 
           ResultMeasureValue, ResultMeasure.MeasureUnitCode, 
           DetectionQuantitationLimitMeasure.MeasureValue)

 # Replace NA values and prepare the data for plotting
  processed_data <- filtered_data |>
    mutate(ResultMeasureValue = as.numeric(coalesce(ResultMeasureValue, 
        as.character(DetectionQuantitationLimitMeasure.MeasureValue)))) |>
    group_by(ActivityStartDate) |>
    slice(1) |> 
    ungroup() |>
    mutate(Year = as.factor(year(ActivityStartDate)),
           OrdinalDate = yday(ActivityStartDate),
           SiteName = site_name)

  # Specify a clear color palette
  color_palette <- c("2006" = "red", "2007" = "blue", "2008" = "green", 
                     "2009" = "orange", "2010" = "purple", "2011" = "cyan", 
                     "2012" = "magenta", "2013" = "brown", "2014" = "pink", 
                     "2015" = "darkgreen", "2016" = "darkblue", "2017" = "darkred", 
                     "2018" = "gold", "2019" = "violet", "2020" = "lightblue", 
                     "2021" = "lightgreen", "2022" = "darkorange", "2023" = "darkviolet")
  
  # Create the plot with OrdinalDate as the x-axis
  plot <- ggplot(processed_data, aes(x = OrdinalDate, y = ResultMeasureValue, group = Year, color = Year)) +
    geom_line(size = 1) +  # Adjust line thickness for better visibility
    geom_point(size = 1) +  # Adjust point size for better visibility
    labs(title = paste(analyte_name, "Levels by Ordinal Date at", site_name),
         x = "Ordinal Date",
         y = paste(analyte_name, "(", result_fraction, "mg/L)", sep = ""),
         color = "Year") +
    theme_minimal() +
    scale_color_manual(values = color_palette) +  # Apply the clear color palette
    scale_x_continuous(breaks = seq(1, 366, by = 30),  # Adjust x-axis ticks
                       labels = function(x) {
                         month.name[x %/% 30 + 1]
                       })

  # Return both processed data and plot
  return(list(data = processed_data, plot = plot))
}

# Example usage for Total Arsenic
#arsenic_results <- visualize_analyte_data_annual_trend("SNR1", "Arsenic", "Total")


```


```{r AnnualTrendPlotPerYear}
BHR1_mg_total <- visualize_analyte_data_annual_trend("BHR1", "Magnesium", "Total")
YRCS_na_dissolved <- visualize_analyte_data_annual_trend("YRCS", "Sodium", "Dissolved")

BHR1_mg_total
YRCS_na_dissolved
```

Other things to do:
1. Compare a single analyte across all monitoring sites
2. Compare all analyes across a single monitoring site
3. Develop a single page reference guide that we can use when lab data is received.
     a. Allow GRYN to immediately verify the lab data is within expected parameters.
     b. No longer have to input lab data into excel spreadsheet.


Last modified, 2025-01-30.
Ben LaFrance, Physical Scientist at the Greater Yellowstone Network
e-mail: benjamin_lafrance@nps.gov

End of script.
